Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-05-24 pkg/paper/Paper/draft.Rnw
Rewrote "Data" section. Seasonal adjustment is now presented as more of a completely natural step (as opposed to a fix of problems in the data). The appendix can probably also be omitted.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@98 b679b2f7-6539-45d0-a605-5ddcb976cde9

Ever since the beginning of the European Union, the topic of a common currency was a controversial issue. Although the Economic and Monetary Union (EMU) is now a fact, the discussion about the economic effects of the euro is far from being settled. The controversial topics range from the question of whether or not the eurozone is indeed an optimal currency area (OCA) all the way to the very survival of the euro in light of the budgetary problems of some of its member states. The effects of monetary unions on a number of macroeconomic indicators, with inflation being the most important one in this context, is in the center of an ongoing debate: this regards the question about short-run and steady-state inflation uncertainty (in context to inflation expectations) dealt with in \citet{cap}, or the structural convergence of the inflation rates in EU countries, which is the topic of an investigation by \citet{palomba}.

The way towards the EMU basically consisted of 3 stages: stage I (1990--1994) as a phase of liberalization, stage II (1995-1998) a phase of convergence and stage III (1999-2001) the transition period, which ended with the introduction of the euro as legal tender. A possible fourth stage is the continuing integration of new member states, mostly former communist countries. Any new state that wants to join the EMU is first obliged to fulfill the Maastricht criteria, which, besides setting rules for government debts and interest rates, also requires the participation in the ERM~II (Exchange Rate Mechanism~II) for two years, while the exchange rates towards the euro are not allowed to cross the nominal band.

@@

However, a deeper inquiry into this problem is in need of a method able to capture changes in inflation dynamics. We contribute to this debate by developing a new method for testing for structural breaks in the dynamics of the inflation rates and thereby upon the effect of the EMU on a number of European countries both within and without the eurozone. The framework developed in this paper makes it possible to test for any significant changes of a countries inflation dynamics and thus helps interpreting the effects of the EMU. We focus on changes in the first three moments of these inflation rates, which we assume to be distributed according to a generalized logistic distribution (GL). The GL distribution is a suitable one due to its flexibility to capture nonsymmetry and because of its fatter tails. For this purpose, we develop a new estimation approach building on previous research by \citet{z07}.

@@

We \fixme{Jesus: yearly inflation rates better; don't start with data in paper}
employ seasonally unadjusted HICP series of 21 countries ranging from January
1990 to December 2010 that are obtained from the \citet{oecd}. Countries
included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,
Finland,	France,	Germany,	Greece,	Hungary,	Ireland,
Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia,
Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as
well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta
are not included since they are very small economies.

The countries in this sample can be divided into three different groups. First
the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece,
Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and
Spain).  Second, the EU members not participating in the ERM~II: Czech Republic,
Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a
member of the EU and the ERM~II, but not yet a member of the EMU.

As  shown in a paper by \cite{pltm}, the harmonization of the treatment of sales
prices led to considerable changes in some countries (Germany, Portugal,
Ireland, Luxembourg, Belgium, Italy and Spain); failing to take this into
account may lead to considerable distortions in the estimation. We dealt with
the inclusion of sales periods\footnote{This was demanded by a Commission
regulation, details may be found in: \citet{eurlex} and in the
Appendix~\ref{sec:sales}.} and all the other seasonality in the data by using
the X-12-ARIMA method for desaisonalization, as described in \cite{X-12-ARIMA}
and \cite{Findley+Monsell+Bell:1998}.

>>>>

Ever since the beginning of the European Union, the topic of a common currency
was a controversial issue. Although the Economic and Monetary Union (EMU) is now
a fact, the discussion about the economic effects of the euro is far from being
settled. The controversial topics range from the question of whether or not the
eurozone is indeed an optimal currency area (OCA) \fixme{Z: include a citation for
OCA or is this completely common knowledge?} all the way to the very
survival of the euro in light of the budgetary problems of some of its member
states. The effects of monetary unions on a number of macroeconomic indicators,
with inflation being the most important one in this context, is in the center of
an ongoing debate: this regards the question about short-run and steady-state
inflation uncertainty (in context to inflation expectations) dealt with in
\citet{cap}, or the structural convergence of the inflation rates in EU
countries, which is the topic of an investigation by \citet{palomba}.

The way towards the EMU basically consisted of three stages: stage~I
(1990--1994) as a phase of liberalization, stage~II (1995--1998) a phase of
convergence and stage~III (1999--2001) the transition period, which ended with
the introduction of the euro as legal tender. A possible fourth stage is the
continuing integration of new member states, mostly former communist countries.
Any new state that wants to join the EMU is first obliged to fulfill the
Maastricht criteria, which, besides setting rules for government debts and
interest rates, also requires the participation in the ERM~II (Exchange Rate
Mechanism~II) for two years, while the exchange rates towards the euro are not
allowed to cross the nominal band.

@@

However, a deeper inquiry into this problem is in need of a method able to
capture changes in inflation dynamics. We contribute to this debate by
developing a new method for testing for structural breaks in the dynamics of the
inflation rates and thereby upon the effect of the EMU on a number of European
countries both within and outside the eurozone. The framework developed in this
paper makes it possible to test for any significant changes of a countries
inflation dynamics and thus helps interpreting the effects of the EMU. We focus
on changes in the first three moments of these inflation rates, which we assume
to be distributed according to a generalized logistic distribution (GL). The GL
distribution is a suitable one due to its flexibility to capture nonsymmetry and
because of its fatter tails. For this purpose, we develop a new estimation
approach building on previous research by \citet{z07}.

@@

All empirical analyses are based on seasonally adjusted inflation rates (in percent)
for \Sexpr{ncol(hicps)}~countries
(\Sexpr{gsub("UK", "United Kingdom", gsub("CzechRep", "Czech Rep", paste(colnames(hicps), collapse = ", "), fixed = TRUE), fixed = TRUE)})
in the time period from the early or mid-1990s to the end of 2010.
The original data source are seasonally unadjusted HICP series provided by the
\citet{oecd} from January~1990 (if available) to December~2010, that were transformed
to inflation rates using log-returns and subsequent seasonal adjustment
via X-12-ARIMA \citep{Findley+Monsell+Bell:1998}.

The countries in this sample can be divided into three different groups: (1)~euro
countries -- Austria, Belgium, Estonia, France, Finland, Germany, Greece,
Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia,
Spain; (2)~EU members not participating in the ERM~II -- Czech Republic,
Hungary, Poland, United Kingdom, Sweden; (3)~Denmark which stands on its own as a
member of the EU and the ERM~II, but not yet of the EMU. 
Latvia, Lithuania, Bulgaria, Romania are excluded due to data scarcity.
Cyprus and Malta are not included since they are very small economies.

Seasonal adjustment is necessary due to the harmonization of the treatment
of sales prices\footnote{Inclusion of sales prices in the HICP was demanded by
a Commission regulation \citep[see][for details]{eurlex}.} in the HICP, resulting in
changes of the seasonal inflation patters over time within several countries
and also between the countries. See \cite{pltm} for a detailed discussion.
